---
title: "Treeline change through time"
author: "Joanna Corimanya"
date: "2023-12-08"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import and clean data}
#set working directory
setwd("~/Desktop/KU/Projects/GeographicTreelinePatterns")
# Read in your data
peaks <- read.csv("data/20231208_Averages.csv", header = TRUE)
#in case there are "X" columns at the end
peaks<-peaks[,-c(17:23)]
#125 are missing directions
missing.dirs<-peaks[peaks$dir == "",]
##remove missing directions
peaks<-peaks[peaks$dir!="",]
#remove NA values
peaks<-na.omit(peaks)

```
Now that the data is cleaned, I created a hierarchical nested list which includes all mountains as a seperate element. Within each mountain, there are 1-8 directions. Within each direction, there are 3 groups of years, each of which I have the average and the maximum value (1984-1988, 1998-2002, 2013-2017).

``` {r create nested list}
#Extract year columns
year_columns <- colnames(peaks[,5:10])
#Function to create nested list for a single peak
create_peak_list <- function(peak_data) {
  # Split by direction
  transects <- split(peak_data, peak_data$dir)
  
  #For each transect, split by year and keep only relevant columns
  lapply(transects, function(transect) {
    lapply(year_columns, function(yr) {
      transect_data <- transect[c("wkt_geom", "Peak_ID", "Country", "dir", "dis_m_", "elevation", "long", "lat", "TreesBegin", yr)]
      names(transect_data)[ncol(transect_data)] <- "NDVI"
      return(transect_data)
    })
  })
}

#Create nested list structure for all peaks
nested_peaks <- lapply(split(peaks, peaks$Name), create_peak_list)

#count number of unique combinations
# Initialize a vector to hold counts for each mountain
dfs_per_mountain <- integer(length(nested_peaks))

# Loop over each mountain index
for (i in seq_along(nested_peaks)) {
  dfs_per_mountain[i] <- length(
    rapply(
      nested_peaks[[i]],
      f       = function(x) x,
      classes = "data.frame",
      how     = "list"
    )
  )
}

# Inspect per-mountain counts
print(dfs_per_mountain)

# Sum to get grand total
total_dfs <- sum(dfs_per_mountain)
total_combos <-total_dfs*2
print(total_combos)

###1864 = total while still including Mt. Washington. Without Mt. Washington transect, total is 1848.
```

Now that I have the list of all peaks, directions, and years, I need a function that removes rows unless they are associated with treeline (Either the last 10 points on a transect or all the points from the edge to treeline, depending on how many transects were measured). This is to account for differences in geography, where sometimes there are small bare peaks on the way up to the peak of interest. Restricting to points around treeline eliminates the noise these cases create, increasing the probability of detecting the true relationship.

```{r find treeline and extract stations}
#function to select relevant stations 
select_stations <- function(df) {
  #Sort the dataframe by distance from the peak 
  df <- df[order(df$dis_m_), ]
  
  #Find the index where TreesBegin is 1
  treeline_index <- which(df$TreesBegin == 1)

  
  if (length(treeline_index) > 0) {
    # Extract NDVI at treeline
    treeline_elevation <- df$elevation[treeline_index[1]]
    #Calculate stations after treeline
    stations_after_treeline <- nrow(df) - treeline_index[1] + 1
    stations_to_use <- max(10, 2 * stations_after_treeline + 1)
    
    #Select the relevant stations
    if (stations_to_use >= nrow(df)) {
      selected_df <- df
    } else {
      selected_df <- df[(nrow(df) - stations_to_use + 1):nrow(df), ]
    }
  } else {
    #If no treeline, select last 10 stations
    selected_df <- tail(df, 10)
    treeline_elevation <- NA
  }
  # Add treeline_elevation as a column to the selected dataframe
  selected_df$treeline_elevation <- treeline_elevation
  return(selected_df)
}

#Apply the selection function to each year in each direction in each peak
selected_nested_peaks <- lapply(nested_peaks, function(peak) {
  lapply(peak, function(direction) {
    lapply(direction, select_stations)
  })
})

```
Now each dataframe only contains the points surrounding treeline. At this point, I performed three types of regressions on each data frame (Linear, reciprocal-linear, and reciprocal-quadratic) to see which model best describes the relationship between NDVI and elevation. The three models were compared via AIC for each peak, direction, and year group. I achieved this by building a function and applying it to the nested list.

```{r regression function and application}
perform_regressions <- function(df, peak_name, direction, year, peak_id, longitude, latitude, country, treeline_elevation) {
  if (!is.data.frame(df) || nrow(df) <= 1) {
    return(data.frame(
      Peak = rep(peak_name, 3),
      Direction = rep(direction, 3),
      Year = rep(year, 3),
      Regression_Type = c("Linear", "Reciprocal_Linear", "Reciprocal_Quadratic"),
      Significance = rep(NA, 3),
      Intercept = rep(NA, 3),
      Slope = rep(NA, 3),
      AIC = rep(NA, 3),
      Best_Model = rep(NA, 3),
      Stations_After_Treeline = rep(NA, 3),
      Peak_ID = rep(peak_id, 3),
      Long = rep(longitude, 3),
      Lat = rep(latitude, 3),
      Country = rep(country, 3),
      Treeline_elevation = rep(treeline_elevation, 3)
    ))
  }
  
  # Exclude rows where NDVI is NA or zero
  df <- df[!is.na(df$NDVI) & df$NDVI != 0,]
  
  # If there are not enough valid rows after filtering, return NA for all values
  if (nrow(df) <= 1) {
    return(data.frame(
      Peak = rep(peak_name, 3),
      Direction = rep(direction, 3),
      Year = rep(year, 3),
      Regression_Type = c("Linear", "Reciprocal_Linear", "Reciprocal_Quadratic"),
      Significance = rep(NA, 3),
      Intercept = rep(NA, 3),
      Slope = rep(NA, 3),
      AIC = rep(NA, 3),
      Best_Model = rep(NA, 3),
      Stations_After_Treeline = rep(NA, 3),
      Peak_ID = rep(peak_id, 3),
      Long = rep(longitude, 3),
      Lat = rep(latitude, 3),
      Country = rep(country, 3),
      Treeline_elevation = rep(treeline_elevation, 3)
    ))
  }
  
  
  #Fit the models
  model_linear <- lm(NDVI ~ elevation, data = df)
  model_reciprocal_linear <- lm(1/NDVI ~ elevation, data = df)
  model_reciprocal_quadratic <- lm(1/NDVI ~ I(1/elevation^2) + elevation, data = df)
  
  #Calculate AICs
  aic_linear <- AIC(model_linear)
  aic_reciprocal_linear <- AIC(model_reciprocal_linear)
  aic_reciprocal_quadratic <- AIC(model_reciprocal_quadratic)
  
  #Calculate coefficients and significance for each model
  coef_linear <- coef(model_linear)
  coef_reciprocal_linear <- coef(model_reciprocal_linear)
  coef_reciprocal_quadratic <- coef(model_reciprocal_quadratic)
  
  significance_linear <- summary(model_linear)$coefficients[2, 'Pr(>|t|)']
  significance_reciprocal_linear <- summary(model_reciprocal_linear)$coefficients[2, 'Pr(>|t|)']
  significance_reciprocal_quadratic <- summary(model_reciprocal_quadratic)$coefficients[2, 'Pr(>|t|)']
  
  #Determine the best model by AIC
  aic_values <- c(Linear = aic_linear, Reciprocal_Linear = aic_reciprocal_linear, Reciprocal_Quadratic = aic_reciprocal_quadratic)
  best_model_type <- names(which.min(aic_values))
  
  #Construct and return a dataframe with results for all model types
  results_df <- data.frame(
    Peak = rep(peak_name, 3),
    Direction = rep(direction, 3),
    Year = rep(year, 3),
    Regression_Type = c("Linear", "Reciprocal_Linear", "Reciprocal_Quadratic"),
    Significance = c(significance_linear, significance_reciprocal_linear, significance_reciprocal_quadratic),
    Intercept = c(coef_linear[1], coef_reciprocal_linear[1], coef_reciprocal_quadratic[1]),
    Slope = c(coef_linear[2], coef_reciprocal_linear[2], coef_reciprocal_quadratic[2]),
    AIC = c(aic_linear, aic_reciprocal_linear, aic_reciprocal_quadratic),
    Best_Model = rep(best_model_type, 3),
    Stations_After_Treeline = rep(nrow(df), 3),
    Peak_ID = rep(peak_id, 3),
    Long = rep(longitude, 3),
    Lat = rep(latitude, 3),
    Country = rep(country,3),
    Treeline_elevation = rep(treeline_elevation,3)
  )
  
  return(results_df)
}

all_results <- list()


for (peak_name in names(selected_nested_peaks)) {
  peak_data <- selected_nested_peaks[[peak_name]]
  for (direction in names(peak_data)) {
    direction_data <- peak_data[[direction]]
    for (year in 1:length(direction_data)) {
      year_data <- direction_data[[year]]
      
      
      #Extract Peak_ID, longitude, and latitude 
      peak_id <- unique(year_data$Peak_ID)[1]
      longitude <- unique(year_data$long)[1]
      latitude <- unique(year_data$lat)[1]
      country <- unique(year_data$Country)[1]
      treeline_elevation = unique(year_data$treeline_elevation)[1]
      # Perform regressions and store results
      regression_results <- perform_regressions(year_data, peak_name, direction, 
                                                year_columns[year], peak_id, longitude, 
                                                latitude,country,treeline_elevation )
      all_results <- c(all_results, list(regression_results))
    }
  }
}

#Combine all results into a single dataframe
final_results_df <- do.call(rbind, all_results) #this works but only preserves rows for the best model


write.csv(final_results_df,"Regressions_28Oct.csv")
```
Now the analysis is complete on each dataframe. This is as far as I have gotten with the project. Below, I built a sample plot that illustrates what I aim to do with each year group in each direction and each year. I will also analyze change in intercept and slope. However, we need to filter the data to only relevant relationships, and we have not yet decided how we will filter prior to comparing the models. therefore, I am unable to go further at this time.
```{r plot example graph}
regressions<-read.csv("Regressions_8Dec.csv") #import data
#Create a plot of one model to show example: Abercrombie Mountain
#subset to significant linear models with negative slopes from 1984-1988 only
sig.88.lin.neg<-regressions[regressions$Significance <= 0.05 
                            & regressions$Year == "Avg8488"&
                              regressions$Regression_Type == "Linear"
                            & regressions$Slope <= 0,] 
#get slope and intercept from 2013-2017
#subset to significant linear models with negative slopes from 2013-2017 only
ab.17<-regressions[regressions$Peak == "Abercrombie Mtn." & regressions$Year == "Avg1317"&
                              regressions$Regression_Type == "Linear"
                            & regressions$Direction == "E",] 
intercept<-sig.88.lin.neg[1,7] 
slope<-sig.88.lin.neg[1,8] 
intercept17<-ab.17[1,7] 
slope17<-ab.17[1,8] 
peaks <- read.csv("20231208_Averages.csv", header = TRUE) # to get elevation for graph
range(peaks$elevation)
png("fig.1.png")
curve(intercept +slope*x, from =261,
      to = 5581, col = "darkseagreen4",
      xlab = "Elevation",
      ylab = "NDVI",
      main = "Eastern Abercrombie Mountain Treeline Change",
      lwd = 3)
#compare with 2017 (add to same plot)
curve(intercept17 +slope17*x, from =261,
to = 5581, col = "peru", add = TRUE,
lwd = 3)
#add legend
legend("topright", legend = c("1984-1988", "2013-2017"), col = c("darkseagreen4", "peru"), lty = 1, lwd = 3)
dev.off()
```
